Practical Effects of Integrating Temperature with Strang Split Reactions

### 2021-03-18 strang/paper.tex
word count

Simulations of stellar flows require solving the equations of
hydrodynamics coupled with a nuclear reaction network.  The equations
of hydrodynamics with reacting sources are:
\begin{align}
\frac{\partial \rho}{\partial t} + \nabla \cdot (\rho \Ub) &= 0 \\
\frac{\partial (\rho X_k)}{\partial t} + \nabla \cdot (\rho X_k \Ub) &= \rho \omegadot_k \\
\frac{\partial (\rho \Ub)}{\partial t} + \nabla \cdot (\rho \Ub \Ub) + \nabla p &= 0 \\
\frac{\partial (\rho E)}{\partial t} + \nabla \cdot \left [ (\rho E + p) \Ub \right ] &= \rho \Sdot
\end{align}
where $\rho$ is the density, $\Ub$ is the velocity vector, $X_k$ are the species mass fractions,
$\omegadot_k$ are the species creation rates (from reactions), $p$ is the pressure, $E$ is the
specific total energy, and $\Sdot$ is the nuclear energy generation rate.

When we are reacting, we can look at internal energy
\begin{equation}
\rho \frac{De}{Dt} + p \nabla \cdot \Ub = \rho \Sdot,
\end{equation}
where $e$ is the specific internal energy or alternately, we can evolve the temperature, 
which we can write as
\begin{equation}
\rho c_v \frac{DT}{Dt} = \rho \left (\frac{p}{\rho^2} - e_\rho \right ) \frac{D\rho}{Dt} + \rho \Sdot,
\end{equation}
where $c_v$ is the specific heat capacity at constant volume, $T$ is the temperature and $e_\rho$ is the derivative of 
the specific internal energy with respect to density (this form neglects the change in 
temperature due to composition
changes from reactions, see \citealt{ABNZ:III}, which assumes chemical
equilibrium).

@@

\acknowledgements \castro\ and all of the code used here is freely available at
\url{https://github.com/AMReX-Astro/Castro}.   The work at
Stony Brook was supported by DOE/Office of Nuclear Physics grant
DE-FG02-87ER40317.  This material is based upon work supported by the
U.S. Department of Energy, Office of Science, Office of Advanced
Scientific Computing Research and Office of Nuclear Physics,
Scientific Discovery through Advanced Computing (SciDAC) program under
Award Number DE-SC0017955.  This research was supported by the
Exascale Computing Project (17-SC-20-SC), a collaborative effort of
the U.S. Department of Energy Office of Science and the National
Nuclear Security Administration.

\software{\amrex\ \citep{amrex_joss},
          \castro\ \citep{castro},
          GNU Compiler Collection (\url{https://gcc.gnu.org/}),
          matplotlib (\citealt{Hunter:2007},  \url{http://matplotlib.org/})
          NumPy \citep{numpy,numpy2},
          python (\url{https://www.python.org/}),
          VODE \citep{vode}
         }

>>>>

Simulations of stellar flows require solving the equations of
hydrodynamics coupled with a nuclear reaction network.  The equations
of hydrodynamics with reacting sources are:
\begin{align}
\frac{\partial \rho}{\partial t} + \nabla \cdot (\rho \Ub) &= 0 \\
\frac{\partial (\rho X_k)}{\partial t} + \nabla \cdot (\rho X_k \Ub) &= \rho \omegadot_k \\
\frac{\partial (\rho \Ub)}{\partial t} + \nabla \cdot (\rho \Ub \Ub) + \nabla p &= 0 \\
\frac{\partial (\rho E)}{\partial t} + \nabla \cdot \left [ (\rho E + p) \Ub \right ] &= \rho \Sdot
\end{align}
where $\rho$ is the density, $\Ub$ is the velocity vector, $X_k$ are the species mass fractions
with creation rates $\omegadot_k$ from reactions, $p$ is the pressure, $E$ is the
specific total energy, and $\Sdot$ is the nuclear energy generation rate.

When we are reacting, we can look at internal energy
\begin{equation}
\rho \frac{De}{Dt} + p \nabla \cdot \Ub = \rho \Sdot,
\end{equation}
where $e$ is the specific internal energy or alternately, we can evolve the temperature, $T$,
via
\begin{equation}
\rho c_v \frac{DT}{Dt} = \rho \left (\frac{p}{\rho^2} - \left .\frac{\partial e}{\partial \rho} \right |_T \right ) \frac{D\rho}{Dt} + \rho \Sdot,
\end{equation}
where $c_v = \partial e / \partial T |_\rho$ (this form neglects composition
changes, see \citealt{ABNZ:III}, which assumes chemical
equilibrium).

@@

\acknowledgements \castro\ is freely available at
\url{https://github.com/AMReX-Astro/Castro}.   The work at
Stony Brook was supported by DOE/Office of Nuclear Physics grant
DE-FG02-87ER40317.  This material is based upon work supported by the
U.S. Department of Energy, Office of Science, Office of Advanced
Scientific Computing Research and Office of Nuclear Physics,
Scientific Discovery through Advanced Computing (SciDAC) program under
Award Number DE-SC0017955.  This research was supported by the
Exascale Computing Project (17-SC-20-SC), a collaborative effort of
the U.S. Department of Energy Office of Science and the National
Nuclear Security Administration.

\software{\amrex\ \citep{amrex_joss},
          \castro\ \citep{castro_joss},
          matplotlib \citep{Hunter:2007},
          NumPy \citep{numpy},
          VODE \citep{vode}
         }
